---
title: "Visualising the History of openSenseMap.org"
author: "Norwin Roosen"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    df_print: kable
    fig_height: 5
    fig_width: 7
    toc: yes
  html_document:
    code_folding: hide
    df_print: kable
    theme: lumen
    toc: yes
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{Visualising the History of openSenseMap.org}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> This vignette serves as an example on data wrangling & visualization with
`opensensmapr`, `dplyr` and `ggplot2`.

```{r setup, results='hide', message=FALSE, warning=FALSE}
# required packages:
library(opensensmapr) # data download
library(dplyr)        # data wrangling
library(ggplot2)      # plotting
library(lubridate)    # date arithmetic
library(zoo)          # rollmean()
```

openSenseMap.org has grown quite a bit in the last years; it would be interesting
to see how we got to the current `r osem_counts()$boxes` sensor stations,
split up by various attributes of the boxes.

While `opensensmapr` provides extensive methods of filtering boxes by attributes
on the server, we do the filtering within R to save time and gain flexibility.
So the first step is to retrieve *all the boxes*:

```{r download}
# if you want to see results for a specific subset of boxes,
# just specify a filter such as grouptag='ifgi' here
boxes = osem_boxes()
```

# Plot count of boxes by time {.tabset}
By looking at the `createdAt` attribute of each box we know the exact time a box
was registered.
With this approach we have no information about boxes that were deleted in the
meantime, but that's okay for now.

## ...and exposure
```{r exposure_counts, message=FALSE}
exposure_counts = boxes %>%
  group_by(exposure) %>%
  mutate(count = row_number(createdAt))

exposure_colors = c(indoor = 'red', outdoor = 'lightgreen', mobile = 'blue', unknown = 'darkgrey')
ggplot(exposure_counts, aes(x = createdAt, y = count, colour = exposure)) +
  geom_line() +
  scale_colour_manual(values = exposure_colors) +
  xlab('Registration Date') + ylab('senseBox count')
```

Outdoor boxes are growing *fast*!
We can also see the introduction of `mobile` sensor "stations" in 2017. While
mobile boxes are still few, we can expect a quick rise in 2018 once the new
[senseBox MCU with GPS support is released](https://sensebox.de/blog/2018-03-06-senseBox_MCU).

Let's have a quick summary:
```{r exposure_summary}
exposure_counts %>%
  summarise(
    oldest = min(createdAt),
    newest = max(createdAt),
    count = max(count)
  ) %>%
  arrange(desc(count))
```

## ...and grouptag
We can try to find out where the increases in growth came from, by analysing the 
box count by grouptag.

Caveats: Only a small subset of boxes has a grouptag, and we should assume
that these groups are actually bigger. Also, we can see that grouptag naming is
inconsistent (`Luftdaten`, `luftdaten.info`, ...)

```{r grouptag_counts, message=FALSE}
grouptag_counts = boxes %>%
  group_by(grouptag) %>%
  # only include grouptags with 8 or more members
  filter(length(grouptag) >= 8 && !is.na(grouptag)) %>%
  mutate(count = row_number(createdAt))

# helper for sorting the grouptags by boxcount
sortLvls = function(oldFactor, ascending = TRUE) {
  lvls = table(oldFactor) %>% sort(., decreasing = !ascending) %>% names()
  factor(oldFactor, levels = lvls)
}
grouptag_counts$grouptag = sortLvls(grouptag_counts$grouptag, ascending = FALSE)

ggplot(grouptag_counts, aes(x = createdAt, y = count, colour = grouptag)) +
  geom_line(aes(group = grouptag)) +
  xlab('Registration Date') + ylab('senseBox count')
```

```{r grouptag_summary}
grouptag_counts %>%
  summarise(
    oldest = min(createdAt),
    newest = max(createdAt),
    count = max(count)
  ) %>%
  arrange(desc(count))
```

# Plot rate of growth and inactivity per week
First we group the boxes by `createdAt` into bins of one week:
```{r growthrate_registered, warning=FALSE, message=FALSE, results='hide'}
bins = 'week'
mvavg_bins = 6

growth = boxes %>%
  mutate(week = cut(as.Date(createdAt), breaks = bins)) %>%
  group_by(week) %>%
  summarize(count = length(week)) %>%
  mutate(event = 'registered')
```

We can do the same for `updatedAt`, which informs us about the last change to
a box, including uploaded measurements.
This method of determining inactive boxes is fairly inaccurate and should be
considered an approximation, because we have no information about intermediate
inactive phases.
Also deleted boxes would probably have a big impact here.
```{r growthrate_inactive, warning=FALSE, message=FALSE, results='hide'}
inactive = boxes %>%
  # remove boxes that were updated in the last two days,
  # b/c any box becomes inactive at some point by definition of updatedAt
  filter(updatedAt < now() - days(2)) %>%
  mutate(week = cut(as.Date(updatedAt), breaks = bins)) %>%
  group_by(week) %>%
  summarize(count = length(week)) %>%
  mutate(event = 'inactive')
```

Now we can combine both datasets for plotting:
```{r growthrate, warning=FALSE, message=FALSE, results='hide'}
boxes_by_date = bind_rows(growth, inactive) %>% group_by(event)

ggplot(boxes_by_date, aes(x = as.Date(week), colour = event)) +
  xlab('Time') + ylab(paste('rate per ', bins)) +
  scale_x_date(date_breaks="years", date_labels="%Y") +
  scale_colour_manual(values = c(registered = 'lightgreen', inactive = 'grey')) +
  geom_point(aes(y = count), size = 0.5) +
  # moving average, make first and last value NA (to ensure identical length of vectors)
  geom_line(aes(y = rollmean(count, mvavg_bins, fill = list(NA, NULL, NA))))
```

We see a sudden rise in early 2017, which lines up with the fast growing grouptag `Luftdaten`.
This was enabled by an integration of openSenseMap.org into the firmware of the
air quality monitoring project [luftdaten.info](https://luftdaten.info).
The dips in mid 2017 and early 2018 could possibly be explained by production/delivery issues
of the senseBox hardware, but I have no data on the exact time frames to verify.

# Plot duration of boxes being active {.tabset}
While we are looking at `createdAt` and `updatedAt`, we can also extract the duration of activity
of each box, and look at metrics by exposure and grouptag once more:

## ...by exposure
```{r exposure_duration, message=FALSE}
duration = boxes %>%
  group_by(exposure) %>%
  filter(!is.na(updatedAt)) %>%
  mutate(duration = difftime(updatedAt, createdAt, units='days'))

ggplot(duration, aes(x = exposure, y = duration)) +
  geom_boxplot() +
  coord_flip() + ylab('Duration active in Days')
```

The time of activity averages at only `r round(mean(duration$duration))` days,
though there are boxes with `r round(max(duration$duration))` days of activity,
spanning a large chunk of openSenseMap's existence.

## ...by grouptag
```{r grouptag_duration, message=FALSE}
duration = boxes %>%
  group_by(grouptag) %>%
  # only include grouptags with 8 or more members
  filter(length(grouptag) >= 8 && !is.na(grouptag) && !is.na(updatedAt)) %>%
  mutate(duration = difftime(updatedAt, createdAt, units='days'))
  
ggplot(duration, aes(x = grouptag, y = duration)) +
  geom_boxplot() +
  coord_flip() + ylab('Duration active in Days')

duration %>%
  summarize(
    duration_avg = round(mean(duration)),
    duration_min = round(min(duration)),
    duration_max = round(max(duration)),
    oldest_box = round(max(difftime(now(), createdAt, units='days')))
  ) %>%
  arrange(desc(duration_avg))
```

The time of activity averages at only `r round(mean(duration$duration))` days,
though there are boxes with `r round(max(duration$duration))` days of activity,
spanning a large chunk of openSenseMap's existence.

## ...by year of registration
This is less useful, as older boxes are active for a longer time by definition.
If you have an idea how to compensate for that, please send a [Pull Request][PR]!

```{r year_duration, message=FALSE}
# NOTE: boxes older than 2016 missing due to missing updatedAt in database
duration = boxes %>%
  mutate(year = cut(as.Date(createdAt), breaks = 'year')) %>%
  group_by(year) %>%
  filter(!is.na(updatedAt)) %>%
  mutate(duration = difftime(updatedAt, createdAt, units='days'))

ggplot(duration, aes(x = substr(as.character(year), 0, 4), y = duration)) +
  geom_boxplot() +
  coord_flip() + ylab('Duration active in Days') + xlab('Year of Registration')
```

# More Visualisations
Other visualisations come to mind, and are left as an exercise to the reader.
If you implemented some, feel free to add them to this vignette via a [Pull Request][PR].

* growth by phenomenon
* growth by location -> (interactive) map
* set inactive rate in relation to total box count
* filter timespans with big dips in growth rate, and extrapolate the amount of
  senseBoxes that could be on the platform today, assuming there were no production issues ;)

[PR]: https://github.com/sensebox/opensensmapr/pulls
