---
title: "Analyzing environmental sensor data from openSenseMap.org in R"
author: "Norwin Roosen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing environmental sensor data from openSenseMap.org in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyzing environmental sensor data from openSenseMap.org in R

This package provides data ingestion functions for almost any data stored on the
open data platform <https://opensensemap.org>.
Its main goals are to provide means for:

- big data analysis of the measurements stored on the platform
- sensor metadata analysis (sensor counts, spatial distribution, temporal trends)

> *Please note:* The openSenseMap API is sometimes a bit unstable when streaming
long responses, which results in `curl` complaining about `Unexpected EOF`. This
bug is beeing worked on upstream. Meanwhile you have to retry the request when
this occurs.

### Exploring the dataset
Before we look at actual observations, lets get a grasp of the openSenseMap
datasets' structure.

```{r}
library(magrittr)
library(opensensmapr)

all_sensors = osem_boxes()
summary(all_sensors)
```

This gives a good overview already: As of writing this, there are more than 600
sensor stations, of which ~50% are running. Most of them are placed outdoors and
have around 5 sensors each.
The oldest station is from May 2014, while the latest station was registered a
couple of minutes ago.

Another feature of interest is the spatial distribution of the boxes. `plot()`
can help us out here:

```{r}
plot(all_sensors)
```

Seems like we have to reduce our area of interest to Germany.

But what do these sensor stations actually measure? Lets find out.
`osem_phenomena()` gives us a named list of of the counts of each observed
phenomenon for the given set of sensor stations:

```{r}
phenoms = osem_phenomena(all_sensors)
str(phenoms)
```

Thats quite some noise there, with many phenomena being measured by a single
sensor only, or many duplicated phenomena due to slightly different spellings.
We should clean that up, but for now let's just filter out the noise and find
those phenomena with the high sensor numbers:

```{r}
phenoms[phenoms > 20]
```

Alright, temperature it is! PM2.5 seems to be more interesting to analyze though.

We should check how many sensor stations provide useful data: We want only those
boxes with a PM2.5 sensor, that are placed outdoors and are currently submitting
measurements:

```{r}
pm25_sensors = osem_boxes(
  exposure = 'outdoor',
  date = Sys.time(), # ±4 hours
  phenomenon = 'PM2.5'
)

summary(pm25_sensors)
plot(pm25_sensors)
```

Thats still more than 200 measuring stations, we can work with that.

### Analyzing sensor data
Having analyzed the available data sources, let's finally get some measurements.
We could call `osem_measurements(pm25_sensors)` now, however we are focussing on
a restricted area of interest, the city of Berlin
Luckily we can get the measurements filtered by a bounding box as well:

```{r}
library(sf)
library(lubridate)

# construct a bounding box: 12 kilometers around Berlin
berlin = st_point(c(13.4034, 52.5120)) %>%
  st_sfc(crs = 4326) %>%
  st_transform(3857) %>% # allow setting a buffer in meters
  st_buffer(units::set_units(12, km)) %>%
  st_transform(4326) %>% # the opensensemap expects WGS 84
  st_bbox()

pm25 = osem_measurements(
  berlin,
  phenomenon = 'PM2.5',
  from = now() - days(31), # defaults to 2 days, maximum 31 days
  to = now()
)

str(pm25)
plot(pm25)
```

Now we can get started with actual spatiotemporal data analysis. First plot the
measuring locations:

```{r}
pm25_sf = osem_as_sf(pm25)
plot(st_geometry(pm25_sf))
```

`TODO`

### Monitoring growth of the dataset
We can get the total size of the data set using `osem_counts()`. Lets create a
time series of that.
To do so, we create a function that attaches a timestamp to the data, and adds
the new results to an existing `data.frame`:

```{r}
build_osem_counts_timeseries = function (existing_data) {
  osem_counts() %>%
    list(time = Sys.time()) %>%     # attach a timestamp
    as.data.frame() %>%             # make it a dataframe.
    dplyr::bind_rows(existing_data) # combine with existing data
}
```

Now we can call it once every few minutes, to build the time series...

```{r}
osem_counts_ts = data.frame()
osem_counts_ts = build_osem_counts_timeseries(osem_counts_ts)
osem_counts_ts
```

Once we have some data, we can plot the growth of data set over time:

```{r}
plot(measurements~time, osem_counts_ts)
```

Further analysis: `TODO`

### Outlook

Next iterations of this package could include the following features

- improved utility functions (`plot`, `summary`) for measurements and boxes
- better integration of `sf` for spatial analysis
- better scaling data retrieval functions
    - auto paging for time frames > 31 days
    - API based on <https://archive.opensensemap.org>
