---
title: "Caching openSenseMap Data for Reproducibility"
author: "Norwin Roosen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Caching openSenseMap Data for Reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

It may be useful to download data from openSenseMap only once.
For reproducible results, the data should be saved to disk, and reloaded at a
later point.

This avoids..

- changed results for queries without date parameters,
- unnecessary wait times,
- risk of API changes / API unavailability,
- stress on the openSenseMap-server.

This vignette shows how to use this built in `opensensmapr` feature, and
how to do it yourself in case you want to save to other data formats.

```{r setup, results='hide'}
# this vignette requires:
library(opensensmapr)
library(jsonlite)
library(readr)
```

## Using the opensensmapr Caching Feature
All data retrieval functions of `opensensmapr` have a built in caching feature,
which serializes an API response to disk.
Subsequent identical requests will then return the serialized data instead of making
another request.

To use this feature, just add a path to a directory to the `cache` parameter:
```{r cache}
b = osem_boxes(grouptag = 'ifgi', cache = tempdir())

# the next identical request will hit the cache only!
b = osem_boxes(grouptag = 'ifgi', cache = tempdir())

# requests without the cache parameter will still be performed normally
b = osem_boxes(grouptag = 'ifgi')
```

Looking at the cache directory we can see one file for each request, which is identified through a hash of the request URL:
```{r cachelisting}
list.files(tempdir(), pattern = 'osemcache\\..*\\.rds')
```

You can maintain multiple caches simultaneously which allows to only store data related to a script in the same directory:
```{r cache_custom}
cacheDir = getwd() # current working directory
b = osem_boxes(grouptag = 'ifgi', cache = cacheDir)

# the next identical request will hit the cache only!
b = osem_boxes(grouptag = 'ifgi', cache = cacheDir)
```

To get fresh results again, just call `osem_clear_cache()` for the respective cache:
```{r clearcache, results='hide'}
osem_clear_cache()        # clears default cache
osem_clear_cache(getwd()) # clears a custom cache
```

## Custom (De-) Serialization
If you want to roll your own serialization method to support custom data formats,
here's how:

```{r data, results='hide', eval=FALSE}
# first get our example data:
measurements = osem_measurements('Windgeschwindigkeit')
```

If you are paranoid and worry about `.rds` files not being decodable anymore
in the (distant) future, you could serialize to a plain text format such as JSON.
This of course comes at the cost of storage space and performance.
```{r serialize_json, eval=FALSE}
# serializing senseBoxes to JSON, and loading from file again:
write(jsonlite::serializeJSON(measurements), 'measurements.json')
measurements_from_file = jsonlite::unserializeJSON(readr::read_file('measurements.json'))
class(measurements_from_file)
```

This method also persists the R object metadata (classes, attributes).
If you were to use a serialization method that can't persist object metadata, you
could re-apply it with the following functions:

```{r serialize_attrs, eval=FALSE}
# note the toJSON call instead of serializeJSON
write(jsonlite::toJSON(measurements), 'measurements_bad.json')
measurements_without_attrs = jsonlite::fromJSON('measurements_bad.json')
class(measurements_without_attrs)

measurements_with_attrs = osem_as_measurements(measurements_without_attrs)
class(measurements_with_attrs)
```
The same goes for boxes via `osem_as_sensebox()`.

```{r cleanup, include=FALSE, eval=FALSE}
file.remove('measurements.json', 'measurements_bad.json')
```
